File-based localization of numerical perturbations in data analysis pipelines

### 2020-04-28 paper.tex
Put figures back on 2 pages
Small fixes

%%%% Commands %%%%
\newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
\newcommand{\note}[2]{\color{blue}Note from \textsc{#1}: #2\color{black}}
\newcommand{\reprozip}[0]{ReproZip\xspace}
\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:}#1\color{black}}
\newcommand{\english}[1]{\uwave{#1}}
\newcommand{\toolname}[0]{Spot\xspace}
\newcommand{\flirt}[0]{\texttt{flirt}\xspace}
\newcommand{\fnirt}[0]{\texttt{fnirt}\xspace}

\title{\toolname: file-based localization of numerical perturbations in data analysis pipelines}
\begin{document}

@@

\toolname identifies the components in a pipeline, at the resolution level of a system process, that produce different
results in different execution conditions. First, a directed
bipartite provenance graph is recorded for each pipeline execution, where nodes
represent application processes and files, and edges represent read and
write file accesses (Figure~\ref{fig:provenance-graph}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified and
disambiguated, resulting in a provenance DAG (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Figure~\ref{fig:provenance-dag}). 
DAGs produced in different conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Figure~\ref{fig:labeled-dag}).
 The resulting labeled graph identifies the
non-reproducible processes in the pipeline. 
\note{Ali}{The graphs show all executed processes captured by the \reprozip tool
except those that their binaries are placed in directory \texttt{/usr/bin/} or \texttt{/bin/}.
We excluded these processes for a better representation of the pipeline execution.}

\begin{figure*}
  \begin{subfigure}[t]{0.3\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{figures/p-graph.pdf}
    \caption{Raw provenance graph (\reprozip output), with transient files shown in gray.}
    \label{fig:provenance-graph}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.3\linewidth}
    \centering
    \includegraphics[width=.7\linewidth]{figures/p-graph-dag.pdf}
    \caption{Provenance DAG, with disambiguated transient files.}
    \label{fig:provenance-dag}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.3\linewidth}
      \centering
      \includegraphics[width=.7\linewidth]{figures/p-graph-dag-labelled.pdf}
     \caption{Labeled DAG comparing 2 execution conditions, showing 1 non-reproducible process.}
     \label{fig:labeled-dag} 
  \end{subfigure}
    % \includegraphics[width=.8\columnwidth]{figures/provenance-graphs}
    \caption{Provenance graphs
    created from the example pipeline in
    Listing~\ref{listing:sample-script}.
    Processes are represented with circles, files with rectangles, and read/write accesses 
    with plain edges. 
    For visualization purposes, the process tree is represented too, with
    dashed edges. Every node in the graph is labeled using the name of the executable run by the
    process or the file name. All the processes are the calls to FSL bet,
    stats and maths made in Listing~\ref{listing:sample-script}. Process \texttt{bet2}
    is forked by process \texttt{bet}: it was captured by \reprozip while it did not
    appear in Listing~\ref{listing:sample-script}.
  }
    \label{fig:spot-example}
  \end{figure*}

@@

Figure~\ref{fig:labeled-dag} illustrates a hypothetical incremental
labeling of the example in Listing~\ref{listing:sample-script}. Process
\texttt{73\#bet2} is labeled as non-reproducible (red) as it produces files
with differences. To prevent the propagation of these differences, the
files produced by \texttt{73\#bet2} in Condition 2 are replaced with the
files produced by \texttt{73\#bet2} in Condition 1. Processes
\texttt{75\#fslmaths} and \texttt{76\#fslstats} are then executed and
labeled as reproducible (green) as they produce files without differences.

@@

The HCP developed a set of pre-processing pipelines to process structural,
functional, and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, and particularly
on PreFreeSurfer and FreeSurfer. 
A detailed description of the analyses done by these
pipelines is available in~\cite{glasser2013}. 
In summary, the PreFreeSurfer pipeline consists of the following steps: 
\begin{itemize}
\item Gradient Distortion Correction (DC), 
\item Alignment and Anatomical Average (AAve), T1w(s), T2w(s),  
\item Anterior/Posterior Commissure Alignment (ACPC-A), 
\item Brain Extraction (BExt), 
\item \tristan{Ali, why isn't BBR cross-modal registration here and in the heatmap?
Ali: I found this main sub-pipelines as the separate scripts in PreFreeSurfer github repo, 
btw, we can add it here. 
I only reprsent sub-pipelines in the heatmap that at leats include one red process.}
\item Bias Field Correction (BFC), 
\item Atlas-Registration (AR).
\end{itemize}
And the FreeSurfer pipeline consists of the following:
\begin{itemize}
\item Image downsampling, 
\item T1w image registration, 
\item T1w image segmentation,
\item Surface placement, 
\item Surface registration.
\end{itemize}
The average processing time per subject is approximately 2~hours for
PreFreeSurfer and 8~hours for FreeSurfer. The average output file size is
approximately 2.7~GB for PreFreeSurfer and 4.1~GB for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}
The \reprozip traced files show that approximately PreFreeSurfer opened  
83K files and created 7.7K processes, and FreeSurfer opened 62K files and 
created 4K processes in total for a single execution, in this case corresponding to a single subject.

We randomly selected 20 unprocessed subjects 
from the HCP data release S500 
available in \href{https://db.humanconnectome.org}{the ConnectomDB repository} as a subset of the 1200 Subject Release.
For each subject, available data consisted of 1 or 2 T1w images and 1 or 2 T2w images, with 
$256 \times 320 \times 320$ voxels of size $0.7 \times 0.7 \times 0.7$ mm.
Acquisition protocols and parameters are detailed in~\cite{van2013wu}.

@@

We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2
CentOS versions. Each subject was processed twice on the same operating
system to detect within-OS variability coming from pseudo-random
operations. We ignored execution-specific information such as file path or
timestamps by using FreeSurfer tools \texttt{mri\_diff},
\texttt{mris\_diff}, and \texttt{lta\_diff} that determine if MRI volumes
or transformations differ. To compare segmentations $X$ and $Y$, we used the Dice coefficient defined as follows:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\]

@@

@@

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
The processes identified as non-reproducible were observed in linear registration 
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL \texttt{maths}  (in Anatomical Average). 
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing 
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\begin{minipage}{\linewidth}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps.
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  fraction of subjects for which the corresponding process wasn't reproducible. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects. }
  \label{fig:pfs_heatmap}
\end{minipage}

\vspace*{1cm}

\begin{minipage}{\linewidth}
  \centering
    \includegraphics[width=.6\linewidth]{figures/pfs-labeled.pdf} 
    \caption{A complete provenance graph from the PreFreesurfer pipeline.
    We excluded six more processes in the pipeline for better visualization, including:
\texttt{imtest}, \texttt{imcp}, \texttt{remove\_ext}, \texttt{fslval}, \texttt{avscale}, and \texttt{fslhd}.}
    \label{fig:complete_pfs}
\end{minipage}
\end{figure*}

To illustrate the magnitude of the differences,
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject. Differences appear to be very substantial, in
particular in areas framed in red.
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png} 
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7). An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here}.}
    \label{fig:fnirt_result}
\end{figure}

@@

Results from our experiments provide insights on the reproducibility of
neuroimaging pipelines. They also allow us to evaluate the approach adopted
in \toolname.

@@

However, file-based analyses also have limitations related to the
granularity at which they operate. Indeed, differences can only be
identified at the level of an entire operating-system process, which can
correspond to arbitrary amounts of code. Narrowing down the analysis to
particular libraries, functions, or even code sections would require
another approach. Similarly, \toolname would not be able to detect
differences in data not saved in files, and passed to subsequent processes
in memory. A common scenario in neuroimaging pipelines is that tools return
results in their standard output, which is parsed by the calling process
and passed to subsequent ones through variables. Differences that occur in data 
that is processed in this way would not be detected by \toolname.

@@

We presented \toolname, a tool to detect the source of numerical
differences in complex pipelines executed in different computational
conditions. \toolname leverages system-call interception through the
\reprozip tool, and therefore can be applied to the most complex pipelines
without requiring their instrumentation. \toolname is available at \url{https://github.com/big-data-lab-team/spot} under MIT license.

>>>>

%%%% Commands %%%%
% \newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
% \newcommand{\note}[2]{\color{blue}Note from \textsc{#1}: #2\color{black}}
\newcommand{\reprozip}[0]{ReproZip\xspace}
\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:}#1\color{black}}
\newcommand{\toolname}[0]{Spot\xspace}
\newcommand{\flirt}[0]{\texttt{flirt}\xspace}
\newcommand{\fnirt}[0]{\texttt{fnirt}\xspace}

\title{File-based localization of numerical perturbations in data analysis pipelines}
\begin{document}

@@

\toolname identifies the components in a pipeline, at the resolution level of a system process, that produce different
results in different execution conditions. First, a directed
bipartite provenance graph is recorded for each pipeline execution, where nodes
represent application processes and files, and edges represent read and
write file accesses (Figure~\ref{fig:provenance-graph}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified and
disambiguated, resulting in a provenance DAG (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Figure~\ref{fig:provenance-dag}). 
DAGs produced in different conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Figure~\ref{fig:labeled-dag}).
 The resulting labeled graph identifies the
non-reproducible processes in the pipeline.

\begin{figure*}
  \begin{subfigure}[t]{0.3\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{figures/p-graph.pdf}
    \caption{Raw provenance graph (\reprozip output), with transient files shown in gray boxes.}
    \label{fig:provenance-graph}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.3\linewidth}
    \centering
    \includegraphics[width=.7\linewidth]{figures/p-graph-dag.pdf}
    \caption{Provenance DAG, with disambiguated transient files.}
    \label{fig:provenance-dag}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.3\linewidth}
      \centering
      \includegraphics[width=.7\linewidth]{figures/p-graph-dag-labelled.pdf}
     \caption{Labeled DAG comparing 2 execution conditions, showing 1 non-reproducible process.}
     \label{fig:labeled-dag} 
  \end{subfigure}
    % \includegraphics[width=.8\columnwidth]{figures/provenance-graphs}
    \caption{Provenance graphs created from the example pipeline in
    Listing~\ref{listing:sample-script}. Processes are represented with
    circles, files with rectangles, and read/write accesses with plain edges. For convenience, the process tree 
    is also shown, with gray dashed edges. Processes forked by
    \texttt{bet} were captured by \reprozip while they did not appear in
    Listing~\ref{listing:sample-script}. Processed associated with
    executables located in \texttt{/usr/bin/} or \texttt{/bin/} are not shown.
  }
    \label{fig:spot-example}
  \end{figure*}

@@

Figure~\ref{fig:labeled-dag} illustrates a hypothetical incremental
labeling of the example in Listing~\ref{listing:sample-script}. Process
\texttt{bet2} is labeled as non-reproducible (red) as it produces files
with differences. To prevent the propagation of these differences, the
files produced by \texttt{bet2} in Condition 2 are replaced with the
files produced by \texttt{bet2} in Condition 1. Processes
\texttt{fslmaths} and \texttt{fslstats} are then executed and
labeled as reproducible (green) as they produce files without differences.

@@

The HCP developed a set of pre-processing pipelines to process structural,
functional, and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, and particularly
on PreFreeSurfer and FreeSurfer. 
A detailed description of the analyses done by these
pipelines is available in~\cite{glasser2013}. 
In summary, the PreFreeSurfer pipeline consists of the following steps: 
\begin{itemize}
\item Gradient Distortion Correction (DC), 
\item Alignment and Anatomical Average (AAve), T1w(s), T2w(s),  
\item Anterior/Posterior Commissure Alignment (ACPC-A), 
\item Brain Extraction (BExt), 
\item Bias Field Correction (BFC), 
\item Atlas-Registration (AR).
\end{itemize}
And the FreeSurfer pipeline consists of the following:
\begin{itemize}
\item Image downsampling, 
\item T1w image registration, 
\item T1w image segmentation,
\item Surface placement, 
\item Surface registration.
\end{itemize}

We randomly selected 20 unprocessed subjects from the HCP data release S500
available in \href{https://db.humanconnectome.org}{the ConnectomDB
repository} as a subset of the 1200 Subject Release. For each subject,
available data consisted of 1 or 2 T1-weighted images and 1 or 2
T2-weighted images, with $256 \times 320 \times 320$ voxels of size $0.7
\times 0.7 \times 0.7$ mm. Acquisition protocols and parameters are
detailed in~\cite{van2013wu}.

@@

We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2
CentOS versions. Each subject was processed twice on the same operating
system to detect within-OS variability coming from pseudo-random
operations. We compared pipeline results using  FreeSurfer tools \texttt{mri\_diff},
\texttt{mris\_diff}, and \texttt{lta\_diff}, to ignore execution-specific information such as file path or
timestamps. To compare segmentations $X$ and $Y$, we used the Dice coefficient defined as follows:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\]

@@

The average processing time per subject was approximately 2~hours for
PreFreeSurfer and 8~hours for FreeSurfer. The average output file size was
2.7~GB for PreFreeSurfer and 4.1~GB for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}
For each subject, PreFreeSurfer accessed
83K files and created 7.7K processes, and FreeSurfer accessed 62K files and 
created 4K processes.

@@

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
The processes identified as non-reproducible were observed in linear registration 
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL \texttt{maths}  (in Anatomical Average). 
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing 
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps.
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  fraction of subjects for which the corresponding process wasn't reproducible. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects. }
  \label{fig:pfs_heatmap}
\end{figure*}

\begin{figure*} 
   \centering
    \includegraphics[width=\linewidth]{figures/pfs-labeled.pdf} 
    \caption{A complete provenance graph from the PreFreesurfer pipeline. For better visualization, processes associated with commands 
    in \texttt{/bin} or \texttt{/usr/bin} were ommitted, as well as 
  \texttt{imtest}, \texttt{imcp}, \texttt{remove\_ext}, \texttt{fslval}, \texttt{avscale}, and \texttt{fslhd}.}
    \label{fig:complete_pfs}
\end{figure*}

To illustrate the magnitude of the differences,
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject. Differences appear to be important, in
particular in the areas framed in red.
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{figures/t2w_alignment.png} 
    \caption{Differences between T2 \fnirt results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7). An animated version of the comparison is available \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/pfs_t2w_alignment.gif}{here}.}
    \label{fig:fnirt_result}
\end{figure}

@@

Our results provide insights on the reproducibility of
neuroimaging pipelines, and on the relevance of the approach implemented
in \toolname for reproducibility studies.

@@

File-based analyses also have limitations related to the
granularity at which they operate. Indeed, differences can only be
identified at the level of an entire operating-system process, which can
correspond to arbitrary amounts of code. Narrowing down the analysis to
particular libraries, functions, or even code sections would require
another approach. Similarly, \toolname would not be able to detect
differences in data not saved in files but instead passed to subsequent processes
in memory. A common scenario in neuroimaging pipelines is that tools return
results in their standard output, which is parsed by the calling process
and passed to subsequent ones through variables.

@@

We presented \toolname, a tool to detect the source of numerical
differences in complex pipelines executed in different computational
conditions. \toolname leverages system-call interception through the
\reprozip tool, and therefore can be applied to the most complex pipelines
without requiring their instrumentation. It is available at
\url{https://github.com/big-data-lab-team/spot} under MIT license.
